---
title: "Capital Bikeshare Competition Analysis"
author: "Joshua Garties"
date: "2025-03-26"
output: html_document
---
```{r set global options, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(include=TRUE, message=FALSE, warning=FALSE)
```

## Introduction
The purpose of this case study is to provide recommendations to Capital Bikeshare on competing with the privately-owned micromobility providers operating in the region. I use publicly available data from [Capital Bikeshare](https://capitalbikeshare.com/system-data) for its system and [Ride Report](https://public.ridereport.com/regions/washington) for micromobility providers.

The slide deck in this repository presents my findings and recommendations. This README and the R Markdown file walk through the steps I took to perform my analysis.

## Background
**Capital Bikeshare** is a bikesharing service that operates across the Washington, DC region. It is owned by the local governments in its service area, who contract its operations out to a private company. Riders typically unlock and return bikes at docking stations installed throughout the region, although they may end e-bike rides without returning them to a dock for an additional fee.

*Key dates:*

* 2010: Capital Bikeshare opens.
* 2018: Capital Bikeshare begins introducing e-bikes.
* 2023: Capital Bikeshare introduces new e-bikes with greater performance and reliability.

Privately-owned **micromobility** companies operate dockless e-bike and scooter service in DC, Arlington, and the University of Maryland. Riders can start and end dockless rideables anywhere, with some limitations on where they can park. For simplicity, I use the term "micromobility" in this case study to refer to these companies' services.

*Key dates:*

* 2017: DC begins permitting micromobility services.
* 2020: DC raises its cap on micromobility rideables to 5,000 e-bikes and 10,000 scooters.
* 2023: DC raises its cap on micromobility rideables to 10,000 e-bikes and 20,000 scooters.

## Load Data
First, install the packages and load the libraries we will be using. 

```{r packages}
# Install packages and load libraries
install.packages("tidyverse",repos = "http://cran.us.r-project.org")
```

```{r libraries}
library("tidyverse")
library(scales)
library(readxl)
```

### Capital Bikeshare Data
Capital Bikeshare provides records for each ride taken. Download CSV files from [Capital Bikeshare's open data site](https://capitalbikeshare.com/system-data) for each of the months in scope and extract them into a single directory. For this case study, I used data from July 2020 to December 2024.

Use this code to combine the CSVs into the dataframe 'cabi_data_raw'.

```{r read cabi data}
# Set the path to your directory
path <- "./cabi_data"

# List all CSV files in the directory
file_list <- list.files(path = path, pattern = "*.csv", full.names = TRUE)

# Read and combine all CSV files, selecting only desired columns
cabi_data_raw <- file_list %>%
  lapply(function(file) {
    df <- read_csv(file)
    df <- select(df, rideable_type, started_at, ended_at)
    return(df)
  }) %>%
  bind_rows()

# Remove file_list and temporary dataframe to save memory
rm(file_list)

head(cabi_data_raw)

# OPTIONAL: Write the combined data frame to a new CSV file to avoid having to perform these steps again
#write_csv(cabi_data_raw, "cabi_202007_202412.csv", col_names = TRUE)
```

### Micromobility Data
Download CSV files from [Ride Report's DC Region page](https://public.ridereport.com/regions/washington). Save a separate CSV for each of the three regions and each rideable type (e-bikes and scooters) for a total of six CSVs. Save them in another directory.

Since these files have relatively few rows, I chose to combine them in Excel. I added the following columns:
* 'region' populated with "DC", "Arlington", and "UMD" according to their source.
* 'rideable_type' populated with "electric_bike" and "scooter" according to their source.

## Clean and Transform Data
### Capital Bikeshare
First we will check for blank/NA values in the data.

```{r cabi NA values}
# Check for NA values
colSums(is.na(cabi_data_raw))
```

Seeing none, we will create new columns to help us manipulate the data.

The new date columns will help us summarize by year, quarter, and month. We need the quarter because the micromobility dataset is summarized by quarter.

The 'duration_mins' column uses the 'started_at' and 'ended_at' datetimes to show how long the ride was. 

```{r transform cabi data}
# create new columns needed for analysis
cabi_data <- cabi_data_raw %>% 
  mutate(year = year(started_at)) %>% 
  mutate(year_month = ym(format(started_at, "%Y-%m"))) %>% 
  mutate(quarter_start = make_date(year(year_month),(quarter(year_month) - 1) * 3 + 1, 1)) %>% 
  mutate(duration_mins = as.numeric(difftime(ended_at, started_at, unit = "mins"))) %>% 
  select(year, quarter_start, year_month, started_at, rideable_type, duration_mins)

head(cabi_data)
```

Now any rows that still have a negative 'duration_mins' value would represent rides where the start_time is after the end_time. These are presumably errors, which we will remove.

```{r filter rows}
# Check the number of rows with negative duration
num_negative_duration <- sum(cabi_data$duration_mins < 0)
num_negative_duration

# Remove affected rows
cabi_data <- cabi_data %>% filter(duration_mins > 0)
```

Next, inspect the dataframe for outliers.

```{r summary statistics}
# Calculate summary statistics for the dataframe
summary(cabi_data$duration_mins)
```

```{r highest and lowest values}
# View the highest and lowest duration_mins values
tail(sort(cabi_data$duration_mins))

head(sort(cabi_data$duration_mins))
```

This shows that there are extremely high 'duration_mins' values, with the maximum thousands of minutes. There are also very low values, under one second.

Extremely high values likely represent bikes that riders lost or did not properly dock, while low values may represent riders accidentally or deliberately ending their ride immediately after they started.

Since there is no natural cutoff for the high values (maybe some people really do decide to rent a bike for several days?), we can remove values that are three standard deviations above and below the mean.

```{r remove by standard deviation}
# Remove values three standard deviations above and below the mean
cabi_data <- cabi_data %>% 
  filter(duration_mins < (mean(duration_mins) + (3 * sd(duration_mins))) & duration_mins > (mean(duration_mins) - (3 * sd(duration_mins))))
```

However, this does not address the low values. We can remove values under one minute, assuming these do not represent actual rides.

```{r remove low values}
# Remove rides < 1 min
cabi_data <- cabi_data %>% 
  filter(duration_mins > 1)
```

Looking again at the summary statistics, we see that the maximum value is still high. We will be using the mean for this analysis, which will help control for the high outliers.

```{r summary statistics, pt. 2}
# Calculate summary statistics for the dataframe
summary(cabi_data$duration_mins)
```

Next, we want to make sure that the 'rideable_type' coding is consistent, since we will be analyzing trends in the use of e-bikes and classic bikes.

```{r identify rideable_type values}
# Show the different values for rideable_type
unique(cabi_data$rideable_type)
```

This reveals three unique values: "docked_bike", "classic_bike", and "electric_bike" when we are only expecting two. I suspect that "docked_bike" and "classic_bike" refer to the same type of bike. "classic_bike" appeared in head(cabi_data_raw) associated with a relatively recent ride, so perhaps "docked_bike" is an old code that was phased out. 

The two operations below will test this theory.

```{r examine rideable_type values}
# Show when rideable_type codes were first and last used.
cabi_data %>% 
  group_by(rideable_type) %>% 
  summarize(first = min(started_at), last = max(started_at))

# Create a bar plot to show how usage changed over time.
ggplot(cabi_data) +
  geom_bar(mapping=aes(x=year_month, fill=rideable_type))
```

The results support the theory that 'docked_bike' was phased out in favor of 'classic_bike'. To keep these codes consistent, recode 'docked_bike' in the dataframe.

```{r replace docked_bike}
# Replace docked_bike with classic_bike to keep variable values consistent
cabi_data$rideable_type[cabi_data$rideable_type == "docked_bike"] <- "classic_bike"
```

Finally, we will create several summary dataframes that we will use for our visualizations.

```{r cabi smmary dataframes}
# Create a dataframe with total rides per year, by rideable_type
cabi_yearly_rideable <- cabi_data %>% 
  group_by(year = year(year_month), rideable_type) %>% 
  summarize(num_rides = n())

# Create a dataframe with total rides per quarter
cabi_quarterly_total <- cabi_data %>% 
  group_by(quarter_start) %>% 
  summarize(num_rides = n())

# Create a dataframe with total rides per quarter, by rideable_type
cabi_quarterly_rideable <- cabi_data %>% 
  group_by(quarter_start, rideable_type) %>% 
  summarize(num_rides = n())
```

### Micromobility
The micromobility data is already summarized at a high level. Manually reviewing it does not reveal any obvious issues, so we will accept Ride Report's cleaning of the underlying data as sufficient. 

First, save the manually prepared Excel file as a dataframen using .name_repair to address spaces in the column headers.

```{r read micromobility data}
# Read micromobility data from Excel
mm_data_raw <- read_xlsx("dmv_micromobility.xlsx", .name_repair = "universal")
head(mm_data_raw)
```

Then, mutate the columns to match the date columns created for the Capital Bikeshare data.

```{r mutate micromobility data}
# Create new date columns
mm_data <- mm_data_raw %>%
  mutate(
    # Extract the year and convert to integer
    year = as.integer(str_extract(Time.Period, "^\\d{4}")),
    # Extract the quarter number (the digit following "Q")
    quarter = as.integer(str_extract(Time.Period, "(?<=Q)\\d")),
    # Calculate the starting month: Q1 -> 1, Q2 -> 4, Q3 -> 7, Q4 -> 10
    month = (quarter - 1) * 3 + 1,
    # Create a Date object for the first day of that month
    quarter_start = make_date(year, month, 1)
  )

head(mm_data)
```

Finally, create summary dataframes similar to the dataframes we created for Capital Bikeshare.

```{r mm summary dataframes}
# Create a dataframe with total rides per year, by rideable_type
mm_yearly_rideable <- mm_data %>% 
  group_by(year, rideable_type) %>% 
  summarize(num_rides = sum(Total.MDS.Trips))

# Create a dataframe with total rides per quarter
mm_quarterly_total <- mm_data %>% 
  group_by(quarter_start) %>% 
  summarize(num_rides = sum(Total.MDS.Trips))

# Create a dataframe with total rides per quarter, by rideable_type
mm_quarterly_rideable <- mm_data %>% 
  group_by(year, quarter_start, rideable_type) %>% 
  summarize(num_rides = sum(Total.MDS.Trips))
```

## Visualize Data
### Capital Bikeshare

The first visual will be a stacked area chart that shows Capital Bikeshare ridership by quarter. While data is available by month, the Ride Report dataset is summarized by quarter, so we will need to summarize Capital Bikeshare by quarter to compare the two datasets.

The chart shows Capital Bikeshare's total quarterly ridership, with the filled regions representing how much ridership came from classic bikes and e-bikes.

Adding a vertical line at March 20, 2023 shows when Capital Bikeshare began introducing the new e-bikes, allowing the reader to see how the introduction affected ridership.

*Takeaway:* Ridership has grown steadily over the past four years. E-bikes appear to account for a larger share of ridership since the new e-bikes were introduced. 

```{r cabi stacked area chart}
# Create stacked area chart
ggplot(data = cabi_quarterly_rideable) +
  geom_area(mapping = aes(x = quarter_start, y = num_rides, fill = rideable_type)) +
    scale_fill_manual(
    name = NULL,
    values = c("classic_bike" = "#F8766D", "electric_bike" = "#00BFC4"),
    labels = c("classic_bike" = "Classic Bike", "electric_bike" = "Electric Bike")
  ) +
  theme(axis.text.y = element_text(angle = 45), legend.position = c(.125, .825), plot.subtitle = element_text(face = "italic")) +
  scale_y_continuous(
    labels = comma) +
  scale_x_date(
    date_breaks = "1 year",
    date_minor_breaks = "3 months",
    date_labels = "%Y"
  ) +
  labs(title = "Capital Bikeshare Ridership",
       subtitle = "Quarterly",
       caption = "Data from Capital Bikeshare",
       x=NULL,
       y = "Rides") +

# Add a vertical line
  geom_vline(xintercept = as_date("2023-03-20"), linetype = "dashed", color = "#619CFF", size = 1.5) +
  annotate(
    "text", 
    label = "New e-bikes introduced", 
    x = as_date("2022-07-01"), 
    y = 1600000, 
    color = "#619CFF",
    fontface = "bold"
    )

ggsave(filename = "cabi_rides_by_rideable_quarterly.png", width = 7, height = 4)
```

To more clearly show that e-bikes account for a larger proportion of ridership over time, we can plot a stacked bar chart. Each bar represents 100% of the ridership for the year, which shows the change in proportions regardless of total ridership.

*Takeaway:* E-bikes have accounted for an increasing proportion of ridership since 2022, and accounted for the majority of ridership in 2024.

```{r cabi stacked bar}
# Create 100% stacked bar chart
ggplot(data = cabi_yearly_rideable) +
  geom_col(
    mapping = aes(x = year, y = num_rides, fill = rideable_type), 
    position = "fill"
  ) +
  scale_fill_manual(
    name = NULL,
    values = c("classic_bike" = "#F8766D", "electric_bike" = "#00BFC4"),
    labels = c("classic_bike" = "Classic Bike", "electric_bike" = "Electric Bike")
  ) +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = percent) +
  labs(title = "Percent of Rideable Type",
       caption = "Data from Capital Bikeshare",
       x = NULL,
       y = NULL
  )

ggsave(filename = "cabi_rides_annual_ebike.png", width = 7, height = 3)
```

### Micromobility

Plot a stacked area chart similar to the chart for Capital Bikeshare, breaking down ridership by e-bikes and scooters.

*Takeaway:* Ridership has grown steadily over the past four years. Most of the growth is from e-bikes.

```{r}
# Create stacked area chart
ggplot(data = mm_quarterly_rideable) +
  geom_area(mapping = aes(x = quarter_start, y = num_rides, fill = rideable_type)) +
  scale_fill_manual(
    name = NULL,
    values = c("scooter" = "#00BA38", "electric_bike" = "#00BFC4"),
    labels = c("scooter" = "Scooter", "electric_bike" = "Electric Bike")
  ) +
  theme(axis.text.y = element_text(angle = 45), legend.position = c(.125, .825), plot.subtitle = element_text(face = "italic")) +
  coord_cartesian(xlim = as.Date(c("2020-01-01", "2024-10-31"))) +
  scale_y_continuous(labels = comma) +
  scale_x_date(
    date_breaks = "1 year",
    date_minor_breaks = "3 months",
    date_labels = "%Y"
  ) +
  labs(title = "Micromobility Ridership",
       subtitle = "Quarterly",
       caption = "Data from Ride Report for DC, Arlington, and University of Maryland (public.ridereport.com)",
       x=NULL,
       y = "Rides")

ggsave(filename = "mm_by_rideable_total.png", width = 7, height = 4)
```

Plot a 100% stacked bar chart similar to Capital Bikeshare chart, breaking down ridership by e-bikes and scooters. 

*Takeaway:* Scooters dominate ridership, but the proportion of e-bikes is growing.

```{r}
# Create 100% stacked bar chart
ggplot(data = mm_yearly_rideable) +
  geom_col(
    mapping = aes(x = factor(year), y = num_rides, fill = rideable_type), 
    position = "fill"
  ) +
  scale_fill_manual(
    name = NULL,
    values = c("scooter" = "#00BA38", "electric_bike" = "#00BFC4"),
    labels = c("scooter" = "Scooter", "electric_bike" = "Electric Bike")
  ) +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = percent) +
  labs(title = "Percent of Rideable Type",
       caption = "Data from Ride Report for DC, Arlington, and University of Maryland (public.ridereport.com)",
       x = NULL,
       y = NULL
  )

ggsave(filename = "mm_annual_type.png", width = 7, height = 3)
```

### Combined

To compare total Capital Bikeshare and micromobility ridership, plot a chart with one line for each service.

*Takeaway:* Micromobility ridership has been consistently higher than Capital Bikeshare ridership since 2021.

```{r}
# Create line chart
ggplot() +
  geom_line(data = cabi_quarterly_total, aes(x = quarter_start, y = num_rides), color = "#F8766D", size = 1.5) +
  geom_line(data = mm_quarterly_total, aes(x = quarter_start, y = num_rides), color = "#C77CFF", size = 1.5) +
  scale_y_continuous(labels = comma) +
  theme(axis.text.y = element_text(angle = 45), plot.subtitle = element_text(face = "italic")) +
  coord_cartesian(xlim = as.Date(c("2021-01-01", "2024-12-31"))) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  annotate(
    "text", 
    x = c(as_date("2024-10-01"), as_date("2024-09-01")), 
    y = c(1450000, 2400000), 
    label = c("Capital Bikeshare", "Micromobility"),
    color = c("#F8766D", "#C77CFF"),
    fontface = "bold"
    ) +
  scale_x_date(
    date_breaks = "1 year",
    date_minor_breaks = "3 months",
    date_labels = "%Y"
  ) +
  labs(title = "Ridership Comparison",
       subtitle = "Quarterly",
       caption = "Data from Capital Bikeshare and Ride Report for DC, Arlington, and University of Maryland (public.ridereport.com)",
       x=NULL,
       y = "Rides")

ggsave(filename = "ridership_comparison.png", width = 7, height = 3)
```

Next, plot a similar chart showing only e-bike ridership for both services. This requires creating two new dataframes containing only e-bike data before plotting. 

As we did above, we'll add a vertical line at March 20, 2023 showing when Capital Bikeshare began introducing the new e-bikes

*Takeaway:* Micromobility e-bikes overtook Capital Bikeshare e-bikes in 2022, but Capital Bikeshare closed the gap after introducing the new e-bikes.

```{r compare e-bike ridership}
# Create new dataframes for only e-bikes.
cabi_quarterly_ebike <- cabi_quarterly_rideable %>% 
  filter(rideable_type == "electric_bike")

mm_quarterly_ebike <- mm_quarterly_rideable %>% 
  filter(rideable_type == "electric_bike")

# Create line chart
ggplot() +
  geom_line(data = cabi_quarterly_ebike, aes(x = quarter_start, y = num_rides), color = "#F8766D", size = 1.5) +
  geom_line(data = mm_quarterly_ebike, aes(x = quarter_start, y = num_rides), color = "#C77CFF", size = 1.5) +
  scale_y_continuous(
    labels = comma
    ) +
  scale_x_date(
    date_breaks = "1 year",
    date_minor_breaks = "3 months",
    date_labels = "%Y"
  ) +
  theme(axis.text.y = element_text(angle = 45), plot.subtitle = element_text(face = "italic")) +
  coord_cartesian(xlim = as.Date(c("2021-01-01", "2024-11-01"))) +
  annotate(
    "text", 
    x = c(as_date("2021-04-01"), as_date("2021-04-01")), 
    y = c(50000, 250000), 
    label = c("Capital Bikeshare", "Micromobility"),
    color = c("#F8766D", "#C77CFF"),
    fontface = "bold"
  ) +
  labs(title = "E-Bike Ridership Comparison",
       subtitle = "Quarterly",
       caption = "Data from Capital Bikeshare and Ride Report for DC, Arlington, and University of Maryland (public.ridereport.com)",
       x=NULL,
       y = "Rides") +
# Add a vertical line
  geom_vline(xintercept = as_date("2023-03-20"), linetype = "dashed", color = "#619CFF", size = 1.5) +
    annotate(
      "text", 
      label = "New e-bikes introduced", 
      x = as_date("2022-08-01"), 
      y = 850000, 
      color = "#619CFF",
      fontface = "bold"
  )

ggsave(filename = "ridership_comparison_ebike.png", width = 7, height = 3)
```

Plot median trip duration across all Capital Bikeshare and micromobility rideables to assess whether riders select their service and rideable depending on the trip they are taking. We use Q3 2024 data because ridership peaks in Q3 every year, and 2024 is the most recent year available. 

This requires creating two new dataframes for the two services showing their median trip duration in each quarter by rideable type, then joining the two dataframes.

*Takeaway:* Median trip duration is similar across all categories, suggesting that riders do not select service or rideable based on their expected trip time.

```{r median ride time}
# Create dataframe for Capital Bikeshare median trip duration
cabi_quarterly_median <- cabi_data %>% 
  group_by(quarter_start, rideable_type) %>% 
  summarize(median_duration = median(duration_mins)) %>% 
  filter(quarter_start == as_date("2024-07-01")) %>% 
  mutate(operator = "cabi") %>% 
  mutate(rideable_type = if_else(rideable_type == "electric_bike", "cabi_ebike", rideable_type))  

View(cabi_quarterly_median)

# Create dataframe for micromobility median trip duration
mm_quarterly_median <- mm_data %>% 
  group_by(quarter_start, rideable_type) %>% 
  filter(quarter_start == as_date("2024-07-01") & jurisdiction == "DC") %>%
  mutate(median_duration = Median.Trip.Duration..minutes.) %>% 
  mutate(operator = "mm") %>% 
  mutate(rideable_type = if_else(rideable_type == "electric_bike", "mm_ebike", rideable_type)) %>% 
  select(quarter_start, operator, rideable_type, median_duration)

View(mm_quarterly_median)

# Join the two dataframes
combined_quarterly_median <- bind_rows(cabi_quarterly_median, mm_quarterly_median)

View(combined_quarterly_median)

# Create bar chart
ggplot(data = combined_quarterly_median) +
  geom_col(
    mapping = aes(
      x = reorder(rideable_type, median_duration), 
      y = median_duration
      ),
    fill = "#CD9600",
    width = 0.5
    ) +
  theme(plot.subtitle = element_text(face = "italic")) +
  labs(title = "Median Trip Duration",
       subtitle = "Q3 2024",
       caption = "Data from Capital Bikeshare and Ride Report for DC (public.ridereport.com)",
       x="Rideable Type",
       y = "Median Trip Duration (minutes)") +
  scale_y_continuous(
    breaks = c(0, 2, 4, 6, 8, 10)
  ) +
  scale_x_discrete(
    labels = c("Micromobility Scooter", "Micromobility E-Bike", "Capital Bikeshare E-Bike", "Capital Bikeshare Classic Bike")
  )

ggsave(filename = "median_duration.png", width = 7, height = 3)
```

